Fusing Filters with Integer Linear Programming

### 2014-06-17 2014betterfusionforfilters/section/01-Introduction.tex
minor changes

To see the effect of clustering, consider the following program:
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (>   0)  xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}

>>>>

To see the effect of clustering, consider the following program:
\begin{code}
 normalize2 :: Array Int -> (Array Int, Array Int)
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (>   0)  xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}

### 2014-06-17 2014betterfusionforfilters/section/03-SizeInference.tex
minor changes

A size may either be a variable $k$ or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints maybe either be trivially $\true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that $k_xs$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $\true$ terms then yields:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
\\
\vdash & k_{gts} = k_1
        ~\wedge~  k_{xs}  ~= k_2
        ~\wedge~  k_{ys1}  = k_2 
        ~\wedge~  k_{ys2}  = k_2
\end{array}
$$

@@

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer a combinator application. Since each name is bound to at most one combinator, we abuse terminology here slightly and write \emph{combinator $n$} when refering to the combinator occuring in the binding of the name $n$.
The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.

>>>>

A size may either be a variable $k$ or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints may either be trivially $\true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that $k_{xs}$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $\true$ terms then yields:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
\\
\vdash & k_{gts} = k_1
        ~\wedge~  k_{xs}  ~= k_2
        ~\wedge~  k_{ys1}  = k_2 
        ~\wedge~  k_{ys2}  = k_2
\end{array}
$$

@@

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer of a combinator application. Since each name is bound to at most one combinator, we abuse terminology here slightly and write \emph{combinator $n$} when refering to the combinator occuring in the binding of the name $n$.
The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.

### 2014-06-17 2014betterfusionforfilters/section/04-ILP.tex
quickhull
fix c_i in 4.5
More desc to benchmarks
Spacing fix
minor changes

% -----------------------------------------------------------------------------
\section{Integer Linear Programming}
\label{s:ILP}
It is usually possible to cluster a program graph in multiple ways. For example, consider the following simple function:
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us
        sum1 = fold (+) 0   us
        ys   = map  (/ sum) incs
    in  ys
\end{code}

@@

For small programs such as @normalizeInc@ it is possible to naively enumerate all possible clusterings, select just those that are \emph{valid} with respect to fusion preventing edges, and chose the one that maximises a cost metric such as the number of array accesses needed. However, as the program size increases the number of possible clusterings becomes too large to naively enumerate. For example, Pouchet et al~\cite{pouchet2010combined} present a fusion system using the polyhedral model~\cite{pouchet2011polyhedral} and report that some simple numeric programs have over 40,000 possible clusterings, with one particular example having $10^{12}$.

@@

% -----------------------------------------------------------------------------
\subsection{Dependency Graphs}
A dependency graph represents the data dependencies of the program to be fused, and we use it as an intermediate stage when producing linear constraints for the ILP problem. The dependency graph contains enough information to determine the possible clusterings of the input program, while abstracting away from the exact operators used to compute each intermediate array. The rules for producing a dependency graphs are in Figure~\ref{f:DependencyGraph}.

@@

% -----------------------------------------------------------------------------
\subsection{Fusion-preventing Path Optimisation}
\label{s:OptimisedConstraints}
The integer linear program defined in the previous section includes more constraints than strictly necessary to define the valid clusterings. If two nodes have a path between them which includes a fusion preventing edge, then we know up front that they must be placed in different clusters. The following function $possible(a,b)$ determines whether there is any possibility that the two nodes $a$ and $b$ can be fused. Similarly the function $possible'(a, b)$ checks whether there is any possibility that the parents of $a$ and $b$ may be fused.
\begin{tabbing}
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
\\[1ex]
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible'(a,b)$ \\
    \>$=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
\end{tabbing}
With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together. Doing this reduces the total number of constraints, and makes the job of the ILP solver easier. The final formulation of the integer linear program follows.

\begin{tabbing}
MMMMM   \= MMMM \= M \= MMMM \= M \= MMMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j} + \Sigma_{i \in V} N \cdot c_i$  \\
           \> (if $possible(i,j)$)         \\
Subject to \\
           \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            
\\[1ex]
           \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     
\\[1ex]
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $(i,j,\fusionpreventing) \in E$)    
\\[1ex]
           \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $possible(i,j)$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $\neg possible(i,j)$) 
\\[1ex]
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  
\\[1ex]
           \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{j',j}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{i',j'}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$ \\
           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) 
\\[1ex]
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\
MMMMM   \= MMMM \= M \= \kill
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        
\\[1ex]
           \> $N = |V|$
\end{tabbing}

>>>>

% -----------------------------------------------------------------------------
\section{Integer Linear Programming}
\label{s:ILP}
It is usually possible to cluster a program graph in multiple ways. For example, consider the following simple function:
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)     us
        sum1 = fold (+) 0    us
        ys   = map  (/ sum1) incs
    in  ys
\end{code}

@@

For small programs such as @normalizeInc@ it is possible to naively enumerate all possible clusterings, select just those that are \emph{valid} with respect to fusion preventing edges, and choose the one that maximises a cost metric such as the number of array accesses needed. However, as the program size increases the number of possible clusterings becomes too large to naively enumerate. For example, Pouchet et al~\cite{pouchet2010combined} present a fusion system using the polyhedral model~\cite{pouchet2011polyhedral} and report that some simple numeric programs have over 40,000 possible clusterings, with one particular example having $10^{12}$.

@@

% -----------------------------------------------------------------------------
\subsection{Dependency Graphs}
A dependency graph represents the data dependencies of the program to be fused, and we use it as an intermediate stage when producing linear constraints for the ILP problem. The dependency graph contains enough information to determine the possible clusterings of the input program, while abstracting away from the exact operators used to compute each intermediate array. The rules for producing dependency graphs are in Figure~\ref{f:DependencyGraph}.

@@

% -----------------------------------------------------------------------------
\subsection{Fusion-preventing Path Optimisation}
\label{s:OptimisedConstraints}
The integer linear program defined in the previous section includes more constraints than strictly necessary to define the valid clusterings. If two nodes have a path between them which includes a fusion preventing edge, then we know up front that they must be placed in different clusters. The following function $possible(a,b)$ determines whether there is any possibility that the two nodes $a$ and $b$ can be fused. Similarly the function $possible'(a, b)$ checks whether there is any possibility that the parents of $a$ and $b$ may be fused.

\newpage

\begin{tabbing}
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
\\[1ex]
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible'(a,b)$ \\
    \>$=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
\end{tabbing}
With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together. Doing this reduces the total number of constraints, and makes the job of the ILP solver easier. The final formulation of the integer linear program follows.

\begin{tabbing}
MMMMM   \= MMMM \= M \= MMMM \= M \= MMMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j} + \Sigma_{i \in V} N \cdot c_i$  \\
           \> (if $possible(i,j)$)         
\\[1ex]
Subject to \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            
\\[1ex]
           \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     
\\[1ex]
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $(i,j,\fusionpreventing) \in E$)    
\\[1ex]
           \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $(i,j, \fusible) \in E$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $(i,j, \fusionpreventing) \in E$)
\\[1ex]
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  
\\[1ex]
           \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{j',j}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{i',j'}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$ \\
           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) 
\\[1ex]
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) 
\\[1ex]
MMMMM   \= MMMM \= M \= \kill
Where      \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        
\\
           \> $N = |V|$
\end{tabbing}
